\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-30 chapters/bttda.tex
Preface
Preface, ironed out overhangs

We propose to implement this idea as a new supervised feature
extraction tensor method that is a generalization of the aforementioned
\ac{hoda}
algorithm and is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
  \item  We develop a forward model for \ac{hoda} to reconstruct a
	      given input tensor from the extracted features.
      \item This allows us to introduce a state-of-the-art \ac{bci} feature extraction
        method based on the block-term tensor structure, named \acf{bttda}.
      \item We evaluate a \ac{bci} decoder based on \ac{bttda} and its special
        PARAFAC-structured case on decoding tasks for both \ac{erp} and \ac{mi}
        paradigm \ac{bci} datasets.
\end{enumerate*}

@@

\subsection{\Acl{hoda}}
\Acl{hoda}~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products:
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
as visualized in \cref{fig:HODA-backward}.
\begin{figure}[t]
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[A \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by \acf{hoda} applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		\Ac{hoda} finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}

@@

\Ac{hoda} optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
or as the partial \ac{hosvd} of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor:
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices:
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of:
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
The iterative process halts after a fixed number of iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full \ac{hoda} procedure is summarized in \cref{alg:HODA}.
\begin{algorithm}
  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

Since \ac{bttda} is specified above as a forward model, a backward procedure
is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
\ac{bttda} to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in \cref{alg:BTTDA}.
\begin{algorithm}
  \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \cref{eq:HODA-backward} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$:
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward \ac{hoda}
projection:
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
\cref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The \ac{rocauc}, or the accuracy in the case of a balanced multi-class problem,
of classification of extracted feature vectors after feature selection is used as
a cross-validation score.
\begin{algorithm}
  \caption[Greedy model selection procedure.]{Greedy model selection.}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
the highest validation score to determine $B$.

Alternatively, a special case of \ac{bttda} can be constructed using only rank-one
blocks such that the resulting forward model adheres to the PARAFAC structure.
The optimal number of rank-one blocks can be found by truncating as above.
We refer to this strategy as PARAFACDA.

@@

To use HODA as a decoder, it is paired with LDA to classify the
extracted feature (HODA+LDA), with hyperparameters $r_k$.
Similarly, we implemented BTTDA+LDA with the proposed BTTDA feature
extraction with hyperparameters $r_k^{(b)}$ for each block $b$ and the number of blocks
$1\leq B\leq16$.
Additionally, we also introduce PARAFACDA+LDA, which is the special case of
BTTDA+LDA where each $r_k^{(b)}=1$, with $B$ and the feature selection
threshold as only hyperparameters.
Hyperparameters were determined separately for each fold using nested
stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
greedy model selection algorithm in \cref{alg:model-selection}.
For HODA+LDA and the HODA blocks in BTTDA+LDA, we chose
$r=r_1=r_2=\ldots=r_K$ with possible values
$\textstyle{1,2,4,8,\ldots,\max_kD_k}$
to reduce computational cost.
Other HODA and BTTDA hyperparameters were set to
$B_\text{max}=10$, $\varepsilon=\num{1e-8}$ and $I_\text{max}=128$.

@@

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \textcite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PARAFACDA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

@@

\section{Discussion}
\subsection{Contribution}
The results in \cref{tab:erp-score} and \autoref{tab:mi-score} show that
in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
exceeding all comparison methods for the two evaluated datasets, but
for MI datasets, results fall short of those of comparison methods.
Performance on the MI classification task could, however, be greatly influenced
by the exact tensorization method used, i.e., the time-frequency transform in
this case, which is not the main focus of this work.

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score for BTTDA+LDA was also slightly
higher than that of
restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

@@

We have introduced \acf{bttda}, a novel,
tensor-based, supervised dimensionality reduction technique optimized for class
discriminability, which adheres to the block-term tensor structure.
\ac{bttda} is a generalization of \acf{hoda} and can also be
applied as a special sum-of-rank-one tensors PARAFACDA model.
The model is obtained by iteratively fitting \ac{hoda} in a deflation scheme,
leveraging a novel forward modeling step.

>>>>

We propose to implement this idea as a new supervised feature
extraction tensor method that is a generalization of the aforementioned
\ac{hoda}
algorithm and is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
  \item  We develop a forward model for \ac{hoda}. It can reconstruct a
	      given input tensor from the extracted features.
      \item This allows us to introduce a state-of-the-art \ac{bci} feature extraction
        method based on the block-term tensor structure, named \acf{bttda}.
      \item We evaluate a \ac{bci} decoder based on \ac{bttda} and its special
        PARAFAC-structured case on decoding tasks for both \ac{erp} and \ac{mi}
        paradigm \ac{bci} datasets.
\end{enumerate*}

@@

\subsection{\Acl{hoda}}
\Acl{hoda}~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
Let $\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$ be a set of $N$ tensors of order$K$. HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products:
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
as visualized in \cref{fig:HODA-backward}.
\begin{figure}[t]
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[A \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by \acf{hoda} applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		\Ac{hoda} finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}

@@

\Ac{hoda} optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
or as the partial \ac{hosvd} of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor:
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices:
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of:
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
The iterative process halts after a fixed number of iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full \ac{hoda} procedure is summarized in \cref{alg:HODA}.
\begin{algorithm}
  \sffamily
  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{align}
	\left\{\mat{A}^*\right\}
  &= \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2  \\
  &= \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
\end{align}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \sffamily
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

Since \ac{bttda} is specified above as a forward model, a backward procedure
is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
\ac{bttda} to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in \cref{alg:BTTDA}.
\begin{algorithm}
  \sffamily
  \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \cref{eq:HODA-backward} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$:
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward \ac{hoda}
projection:
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
\cref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The \ac{rocauc}, or the accuracy in the case of a balanced multi-class problem,
of classification of extracted feature vectors after feature selection is used as
a cross-validation score.
\begin{algorithm}
  \sffamily
  \caption[Greedy model selection procedure.]{Greedy model selection.}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
the highest validation score to determine $B$.

Alternatively, a special case of \ac{bttda} can be constructed using only rank-one
blocks such that the resulting forward model adheres to the PA\-RA\-FAC structure.
The optimal number of rank-one blocks can be found by truncating as above.
We refer to this strategy as PARAFACDA.

@@

To use HODA as a decoder, it is paired with LDA to classify the
extracted feature (HODA+LDA), with hyperparameters $r_k$.
Similarly, we implemented BTTDA+LDA with the proposed BTTDA feature
extraction with hyperparameters $r_k^{(b)}$ for each block $b$ and the number of blocks
$1\leq B\leq16$.
Additionally, we also introduce PARAFACDA+LDA, which is the special case of
BTTDA+LDA where each $r_k^{(b)}=1$, with $B$ and the feature selection
threshold as only hyperparameters.
Hyperparameters were determined separately for each fold using nested
stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
greedy model selection algorithm in \cref{alg:model-selection}.
For HODA+LDA, as well as for the HODA blocks in BTTDA+LDA, we chose
$r=r_1=r_2=\ldots=r_K$ with possible values
$\textstyle{1,2,4,8,\ldots,\max_kD_k}$
to reduce computational cost.
Other HODA and BTTDA hyperparameters were set to
$B_\text{max}=10$, $\varepsilon=\num{1e-8}$ and $I_\text{max}=128$.

@@

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \textcite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PA\-RA\-FAC\-DA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

@@

\section{Discussion}
\subsection{Contribution}
The results listed in \cref{tab:erp-score} and \autoref{tab:mi-score} show
that, for the tested ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
exceeding all comparison methods for the two evaluated datasets, but
for MI datasets, results fall short of those of comparison methods.
Performance on the MI classification task could, however, be greatly influenced
by the exact tensorization method used, i.e., the time-frequency transform in
this case, which is not the main focus of this work.

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score obtained by BTTDA+LDA was also slightly
higher than that of the
restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

@@

We have introduced \acf{bttda}, a novel,
tensor-based, supervised dimensionality reduction technique optimized for class
discriminability, which adheres to the block-term tensor structure.
\ac{bttda} is a generalization of \acf{hoda} and can also be
applied as a special sum-of-rank-one tensors PA\-RA\-FAC\-DA model.
The model is obtained by iteratively fitting \ac{hoda} in a deflation scheme,
leveraging a novel forward modeling step.

### 2024-11-30 chapters/conclusion.tex
Preface, ironed out overhangs

Coincidentally, advances in gaze-independent decoding also build towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this, and be a valuable addition to a \ac{bci}
assistive technology device.

@@

The optimal approach should unify decoder development, paradigm choice, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several user groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

>>>>

Coincidentally, making advances in gaze-independent decoding also builds towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this, and be a valuable addition to a \ac{bci}
assistive technology device.

@@

The optimal approach should unify decoder development, choice of paradigm, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several user groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

### 2024-11-30 chapters/conflict.tex
Preface, ironed out overhangs

The authors acknowledge the support by the RITMEA project co-financed by the
European Union with the European Regional Development Fund, the French state,
and the Hauts-de-France Region Council.

>>>>

The authors acknowledge the support by the RITMEA project co-fi\-nan\-ced by the
European Union with the European Regional Development Fund, the French state,
and the Hauts-de-France Region Council.
